Kindly Bent to Free Us

### 2019-07-11 kindly/intro.tex
checkpoint

Our work provides a solution to programmers in this situation. We
propose an extended type system for ML-like languages that comes with
linear and affine types, mutable and immutable borrows, but all that
integrated with full type inference, garbage collected data, and
ANOTHER FEATURE.

Compared to the most closely related contender in this design space, Linear Haskell
\cite{DBLP:journals/pacmpl/BernardyBNJS18}, their goals and means are
similar as both systems permit abstraction over the number of uses of
values and retain type inference, but the details are quite different.
\begin{enumerate}
\item Their multiplicities are either linear or unrestricted whereas
  we also permit affine values.
\item Our system supports mutable and immutable borrows, hence the
  need for affinity.
\item Our multiplicities are directly attached to the type of the
  value. For example, in our system the function type \lstinline/'a-{lin}>'b/
  denotes the type of a \emph{single-use function} that can be called
  just once, whereas the multiplicities in Linear Haskell choose
  between $\alpha\to\beta$ and $\alpha \multimap\beta$ where the
  latter is a function that promises to \emph{use its argument exactly once}.
\end{enumerate}
\subsection{First examples}
\label{sec:first-example}
\begin{figure}[tp]
  \begin{subfigure}[t]{0.45\linewidth}
    \lstinputlisting{code/writefiles.affe}
    \caption{File API}
    \label{fig:writing-files-api}
  \end{subfigure}
  ~
  \begin{subfigure}[t]{0.45\linewidth}
\begin{lstlisting}
let main () =
  let h = File.fopen "foo" in
  File.write &!h "Hello ";
  File.write &!h "world!";
  File.close h
\end{lstlisting}
    \caption{File example}
    \label{fig:writing-files-example}
  \end{subfigure}
  \caption{Writing files}
  \label{fig:writing-fiules}
\end{figure}

@@

This example demonstrates two features of our system:
\begin{enumerate}
\item full type inference without annotations in user code,
\item types carry multiplicity annotations in the form of kinds.
\end{enumerate}

@@

\begin{itemize}
\item A type system that encodes linearity and affinity with
  borrowing. The type system is polymorphic with respect to types as
  well as kinds that express constraint on the number of uses of a value.
\item An extension of the HM(X) framework
  \cite{DBLP:journals/tapos/OderskySW99} with kinds, regions, and
  multiplicity constraints that supports full, prinicipal type inference.
\item Constraint simplification techniques.
\item Automatic inference of regions for borrows.
\end{itemize}

>>>>

Our work provides a solution to programmers in this situation. We
propose an extended type system for ML-like languages that comes with
linear and affine types, mutable and immutable borrows, but all that
integrated with full principal type inference, garbage collected data, and
automatic placement of borrowing regions.

The most closely related contenders in this design space are Linear Haskell
\cite{DBLP:journals/pacmpl/BernardyBNJS18}, henceforth called LH,
Quill \cite{DBLP:conf/icfp/Morris16}, and ALMS \cite{DBLP:conf/popl/TovP11}.
Compared to LH, the goals and means are
similar as both systems permit abstraction over the number of uses of
values and retain type inference, but the details are quite different.
\begin{enumerate}
\item LH's multiplicities are either linear or unrestricted whereas
  we also permit affine values.
\item Our system supports mutable and immutable borrows, hence the
  need for affinity.
\item Our multiplicities are directly attached to the type of the
  value. For example, in our system the function type \lstinline/'a-{lin}>'b/
  denotes the type of a \emph{single-use function} that can be called
  just once, whereas the multiplicities in LH choose
  between $\alpha\to\beta$ and $\alpha \multimap\beta$ where the
  latter is a function that promises to \emph{use its argument exactly
    once}.
\item Similar to Quill, we make use of multiplicity contraints. We
  rely on kind subsumption to simplify the inferred types.
\end{enumerate}
See \cref{sec:related-work} for further in-depth discussion of these
and other related works.
\subsection{First examples}
\label{sec:first-example}
\begin{figure}[tp]
  \begin{subfigure}[t]{0.45\linewidth}
    \lstinputlisting{code/writefiles.affe}
    \caption{File API}
    \label{fig:writing-files-api}
  \end{subfigure}
  ~
  \begin{subfigure}[t]{0.45\linewidth}
\begin{lstlisting}
let main () =
  let h = File.fopen "foo" in
  File.write &!h "Hello ";
  File.write &!h "world!";
  File.close h
\end{lstlisting}
    \caption{File example}
    \label{fig:writing-files-example}
  \end{subfigure}
  \caption{Writing files}
  \label{fig:writing-fiules}
\end{figure}

@@

This example demonstrates three features of our system:
\begin{enumerate}
\item full type inference without annotations in user code,
\item types carry multiplicity annotations in the form of kinds,
\item resource APIs can be written in direct style as linearity is a
  property of the type \lstinline/File.t/.
\end{enumerate}

@@

\begin{itemize}
\item A type system that encodes linearity and affinity with
  borrowing. The type system is polymorphic with respect to types as
  well as kinds that express constraints on the number of uses of a value.
\item An extension of the HM(X) framework
  \cite{DBLP:journals/tapos/OderskySW99} with kinds, regions, and
  multiplicity constraints that supports full, prinicipal type
  inference.
\item Full account of the metatheory for type inference and type soundness.
\item Constraint simplification techniques.
\item Automatic inference of regions for borrows.
\item An implementation of the type inference algorithm including all
  constraint simplification and extended with pattern matching. A
  webfront for the algorithm is available, its URL is given in the
  non-anonymous supplementary material, and we plan to submit it for
  artifact evaluation.
\end{itemize}
